---
title: "Advanced_R-Subsetting"
author: "Chris Lavoie"
date: "February 26, 2018"
output: html_document
---

## Subsetting

### Overview

- Three subsetting operators
- Six types of subsetting
- There are multiple differences in behaviour for each object type
- Use subsetting with assignment

### Data types

#### Atomic Vectors

```{r}
x <- c(1.1,2.2,3.3,4.4)

# Positive Integers gives location
# Grabs vector position 1 and 3
x[c(1,3)]

# Grabs item 2
x[2]

# Integer indexs are truncated -> Doesn't evaluate the decimal, only the integer
# Will both return position one
x[c(1.9, 1.1)]

# Negative integers omit that selection
# Returns all but the second position
x[-2]

x[-c(1,4)]

# Cannot mix positive and negative

```

#### Atomic Vectors and Logic

- The most important part of subsetting
- Expressions evaluate to logical operations
- Logical evaluations are applied to each element
```{r}

#
x > 2
# Will return all elements where the expression evaluates to true
x[x > 2]
```
- If logical is shorter than the length of the vector, the logical vector will be recyled until it is equal length

- Subsetting NA always yields NA

```{r}
x[c(TRUE, FALSE, TRUE, NA)]
```
- Character vectors can return elements if the vector is named and the names match
  - Names must match exactly!
```{r}
x <- c(a=1,b=2)

x[1]
x["a"]
#TRUE
x["a"] == x[1]
```
- Name and integer indices can be repeated



### Lists
- Work like vectors
- using [] will return a list.
- using [[]] will return a component of a list
- $ is equivalent to [[]]

```{r}
l <- list(a=1, b=c(1,2), c="fork")

# Returns list of 1
str(l[2])
# Returns element
str(l[[3]])

l["c"]
```


### Matrices and Arrays
- Subset via
  - Using multiple vectors
  - Using a single vector
  - Using a matrix
  
- Most common way is by supplying a 1d index for each dimension

```{r}
m <- matrix(1:100, nrow=10, ncol=10)

m[1,1]
# 2nd row, 6th column
m[2,6]
# Third row, whole column
m[3,]
# Rows 1 2 and 3, columns 2 3 4 and 5
m[1:3, 2:5]

str(m[m > 50])
```

- Using [] will simplify to the lowest possible dimension
- Matrices can also be subsetted like vectors.

```{r}
m[4]

m[c(4,12)]
```

- Higher dimensional arrays can be subset via integer matrices.


### Data frames

- Recall: Possess characteristics of both lists and matrices.
  - If subset with one vector, they behave like a list
  - If subset with two vectors, they behave like a matrix
  
```{r}
df <- data.frame(m)

df["X1"]
df[1:5,"X2"]
# Behaves like a matrix
df[c(1,3),]
# Behaves like a list
df[c(1,3)]
```

- Recall that matrix subsetting simplifies by default
- List subsetting does not!
```{r}
# List subsetting
print("List Subsetting")
str(df["X1"])

# Matrix subsetting
print("Matrix Subsetting")
str(df[,"X1"])
```

### Exercises

1. Fix each of the following common data frame subsetting errors:

mtcars[mtcars$cyl = 4, ]
mtcars[-1:4, ]
mtcars[mtcars$cyl <= 5]
mtcars[mtcars$cyl == 4 | 6, ]
```{r}
mtcars[mtcars$cyl ==4, ] # Fixed equal sign
mtcars[-c(1:4), ] # Made both operations as negative (cannot mix)
mtcars[mtcars$cyl <= 5,] # Added comma
mtcars[mtcars$cyl == 4 || 6, ] # Fixed "or" operator
```

2. Why does x <- 1:5; x[NA] yield five missing values? (Hint: why is it different from x[NA_real_]?)
  - NA is a logical of length one and, similar to a logical vector that is shorter than the vector it is being subset on, it will be recycled to go along the whole vector, unlike if we used NA_real_ (which is a double)
  
```{r}
typeof(NA_real_)
typeof(NA)
```

3. What does upper.tri() return? How does subsetting a matrix with it work? Do we need any additional subsetting rules to describe its behaviour?
- upper.tri() returns a logical matrix of equal dimensions. Any element on the upper triangular of the matrix returns TRUE, and anything else returns FALSE
- Subsetting works like it would for any logical vector in a matrix.
```{r}
x <- outer(1:5, 1:5, FUN = "*")

upper.tri(x)
x[upper.tri(x)]
```
4. Why does mtcars[1:20] return an error? How does it differ from the similar mtcars[1:20, ]?
  - mtcars[1:20] subsets like a list. It tries to return columns 1 - 20, but there are not that many columns in the dataframe
  - mtcars[1:20,] subsets like a matrix. This returns the first twenty rows, and all columns
  
```{r}
mtcars[1:5,]
```


5. Implement your own function that extracts the diagonal entries from a matrix (it should behave like diag(x) where x is a matrix).
```{r}
# returns all entries on the diagonal

# Using the classic loop method
diag_matrix <- function(a_matrix) {
  matrix_rows <- 1:nrow(a_matrix)
  matrix_cols <- 1:ncol(a_matrix)
  
  logical_matrix = matrix(FALSE,nrow = nrow(x), ncol=ncol(x))
  
  for (i in seq_along(matrix_rows)) {
    for (j in seq_along(matrix_cols)) {
      if (i == j) {
        # If the row number is the column number, return true
        logical_matrix[i,j] <-  TRUE
      }
    }
  }
  # use the logical matrix that only returns true
  # on the diagonal to subset the original matrix
  new_matrix <- a_matrix[logical_matrix]
  return(new_matrix)
  

}
f <- diag_matrix(x)
# Look mom I did it
f == diag(x)
```

6. What does df[is.na(df)] <- 0 do? How does it work?
  - Subsets the dataframe by only selecting those who return TRUE to the logical is.na() result.
  - Assigns only the result of the subset to 0!
  
## Subsetting Operators
- Recall: Using [[]] (or $) returns a single value.
- Important when working with lists. Using [] on lists always returns a list
- Because [[]] only returns a single value, can only use with positive integer or single string
- Since dataframes are lists of columns, use [[]] or $ to extract a single column


### Simplifying Vs. Preserving
- Preserving: Get the same type of output as the input
- $ is a shorthand for [[, exact=FALSE]]
  - This means you can use as incomplete strings
  
```{r}
# Extracts mpg
mtcars$m
# Returns null since two columns are matched, which is not alowed with [[]]
mtcars$d
```
- df$x is equivalent to df[["x", exact=FALSE]] - note that the quotations are added.
  - It's best not to mess with this when programming.


### Exercise
1. Given a linear model, e.g., mod <- lm(mpg ~ wt, data = mtcars), extract the residual degrees of freedom. Extract the R squared from the model summary (summary(mod))
```{r}
mod <- lm(mpg ~ wt, data = mtcars)
summ <- summary(mod)

rsquared <- summ$r.squared
rdf <- mod$df.residual

rsquared
rdf
```


## Subset and Assignment
- All subsetting operators can be combined with assignment to modify values
```{r}
x <- 1:5
x[c(1, 2)] <- 2:3
x
```

- There is no checking for duplicates in assignment. Last one wins.
- Cannot combine integers with NA
- But can combine integer with NA. NA is Treated as FALSE

- Empty subsetting can be useful to preserve the structure and class of the object

```{r}
# Keeps the dataframe class
mtcars[] <- lapply(mtcars, as.integer)
class(mtcars)
# Becomes a list
mtcars <- lapply(mtcars, as.integer)
class(mtcars)
```

