---
title: "Advanced R Book Notes"
author: "Chris Lavoie"
date: "February 26, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Advanced R

This is a notebook that will be posted publicly that works through my own personal notes and examples from the Advanced R Textbook by Hadley Wickham, which can be found at http://adv-r.had.co.nz/

```{r test}
print("testing")
```

## Data Structures

### Vector

Three Main Properties
- Type
- Length
- Attributes
Two types
  - Atomic Vectors
      - Come in 6 types: logical, integer, double, character, complex and raw (latter two are rare)
      - Always flat (if nested, will be output as flat)
      - Missing Values are NA


```{r vectors, echo=TRUE}

print("Flat")
print(c(1,2,c(3,4)))


```

#### Vector Types and Tests
 - Use typeof() to determine type, or if specific type is is.*() function
 - All elements of a vector must be same type
  - If different types in the vector, the types will be coeced into most flexible type
    - Most flexible types
      1. Character
      2. Double
      3. Integer
      4. Logical
      
```{r vector coercion}

test_vec <- c(1, "a")

print_and_eval <- function(x) {
  print(x)
  print(typeof(x))
}
# Very clearly we assigned the 1 as an integer, however it was forced into a character vector by coercion
sapply(test_vec, print_and_eval)

```

```{r vector logical}
test_vec <- c(TRUE, FALSE, 1)
test_vec_2 <- c(TRUE, FALSE, 1, "a")

print(test_vec)
print(test_vec_2)


```

Note above that (logical + numeric) = numeric. True is 1 and False is 0
(logical + numeric + character) = character. True is the character "True" and vice versa with false.

- Vector coercion can be taken advantage of in using booleans
```{r operations on booleans}
new_vec <- c(TRUE, FALSE, 1)
# Will sum to 2, as it will be evaluated to c(1,0,1)
sum(new_vec)
```
- Use as.*() whenever necessary to ensure type coercion


### List

- Elements can be of any type, including more lists
```{r}
new_list <- list("a", 1, TRUE, 3.5)
print("Structure of List")
# Each type will be preserved
str(new_list)
print("Structure of Vec")
# Will evaluate to all character, as it is the most flexible type
new_vec <- c("a", 1, TRUE, 3.5)
str(new_vec)
```

- Lists can recursively hold more lists.
  - Vectors can hold lists, but if given a vector of lists and other vectors, will coerce to list and flatten
  
```{r}
vec_of_lists_and_vecs <- c(list(1,2), c(3,4))
str(vec_of_lists_and_vecs)
typeof(vec_of_lists_and_vecs)
```

- Note that the vector will coerce the items in the vector to the most flexible type before evaluating as a list
```{r}
tricky_vec <- c(c(TRUE, 0), c(FALSE, "a"), list("b", 3, TRUE))

str(tricky_vec)
paste("Length:", length(tricky_vec))
print(tricky_vec)
```
- First vector in tricky_vec transforms TRUE to 1, and keeps zero (as integer is the most flexible type)
- Second vector in tricky_vec transforms FALSE to a character and keeps "a"
- The list (element 3) keeps all types
- The entire list evaluated in tricky_vec is flat.

- If we change the tricky_vec into a tricky_list...
```{r}
tricky_list <- list(c(TRUE, 0), c(FALSE, "a"), list("b", 3, TRUE))
str(tricky_list)
# Structure is kept, and not flattened
paste("Length:", length(tricky_list))
```

- can test for list by is.list() command.
- IMPORTANT: Dataframes are lists of lists!


### EXERCISES

1. What are the six types of atomic vector? How does a list differ from an atomic vector?
  - In order of flexibility (least to most): Logical, Integer, Double, Character. Also Complex and Raw
  - Vectors automatically flatten contents and coerce to most flexible type. Lists keep structure and retain type of data.
  
2. What makes is.vector() and is.numeric() fundamentally different to is.list() and is.character()?
  - is.vector() checks if object is a vector with no attributes apart from names. Need to use is.atomic() or is.list()
  - is.numeric() checks for numberliness. Evaluates true for integer and double. Need to specity is.integer() or is.double()
  - is.list() checks if is a list, and only returns TRUE for objects typeof list. Vice versa for character
  
3. Test your knowledge of vector coercion rules by predicting the output of the following uses of c():
```{r}
# c(1, 0)
c(1, FALSE)
#c("a", "1")
c("a", 1)
#list(1, "a")
c(list(1), "a")
#c(1, 1)
c(TRUE, 1L)
```
4. Why do you need to use unlist() to convert a list to an atomic vector? Why doesn’t as.vector() work?
  - list() is a type of vector. You are converting from one type of vector (list) to another (atomic)
  
  
5. Why is 1 == "1" true? Why is -1 < FALSE true? Why is "one" < 2 false?
  - In the evaluation, both elements are converted to the most flexible form of the two.
  
```{r}
1 == "1" # Both are converted to string

-1 < FALSE # Both are converted to double. 0 is greater than -1

"one" < 2 # Both are converted to character. the letters seem to come before numbers.
```


6. Why is the default missing value, NA, a logical vector? What’s special about logical vectors? (Hint: think about c(FALSE, NA_character_).)
  - Logical vectors are Reserved Words that can be coerced into any type. TRUE and FALSE and NA are reserved words. This allows it to work with any type of vector
  

## Attributes

- All objects have them
- Stores metadata into the object
- Can be thought of as a named list

```{r}
y <- 1:10
str(y)
attr(y, "hey_an_attribute") <- "This is something"

attr(y, "hey_an_attribute")

attributes(y)
# Returns the type of the attribute value
typeof(attr(y, "hey_an_attribute"))
# Returns a named list
typeof(attributes(y))
```

- By default, all attributes are lost in manipulation except the three most important (which are not accessed by attr(object, [attribute]))
  - Names: names(x)
    - Can name upon creation: c(a=1, b=2, c=3) or modify: x <- c(1,2,3); names(x) <- c("1", "2", "3"). And a third annoying way
    - Don't have to be unique, but they should.
    - Not all items in a vector need a name. If names attribute is non-empty, missing names will return NA_character_. If empty, returns NULL
  - Dimensions: dim(x)
  - Class: class(x)
  
- An important use of attributes is in the Factor variable.

### Factors
- Factor is a vector that can only contain predefined values
- Built on top of the class factor(), which allows the vector to behave differently
- Uses the levels() attribute, which is also not assignable.
  - levels defines the set of allowable values in the vector
  
- Factors are useful when you know beforehand the set of allowable values
   - the properties of a factor are useful in analysis.
   
```{r}
factor_var <- as.factor(c("a", "b", "c", "a", "a", "b"))
levels(factor_var) <- c("a", "b", "c", "d")
# Here, it is plainly obvious that one of the possible classes is missing.
# This can be helpful for those who are reading your analysis without too much background
table(factor_var)
```


- Factors can be automatically created when reading in CSVs. For this reason, you want to be careful!
  - This is especially true for character vectors
- Use class([VAR]) to figure out if it is a vector or not. You'll be glad you did!

- Note: A lot of the wrongdoings of factors in reading in data is undone with the readr package

### Exercises

1. An early draft used this code to illustrate structure():
```{r}
structure(1:5, comment = "my attribute")
```
But when you print that object you don't see the comment attribute. Why? Is the attribute missing, or is there something else special about it? (Hint: try using help.)

- A comment is a special attribute that is not printed. But if we assigned the comment attribute to an object, we could query it
```{r}
f <- structure(1:5, comment = "my attribute")
# Not printed
f
# queried using a special function
comment(f)
```

2. What happens to a factor when you modify its levels?
```{r}
f1 <- factor(letters) # levels is a default attribute
f1
levels(f1) <- rev(levels(f1)) # rev -> reverse

f1

```
- The letters reversed. The levels assign an order, and the printing of the factor respects that order.

3. What does this code do? How do f2 and f3 differ from f1? 
```{r}
f2 <- rev(factor(letters))
f2
f3 <- factor(letters, levels = rev(letters))
f3
```

- The levels stay the same (as they were never changed), but the order of the initial vector changed.
- f3 is like f2 but inverted. In either case, the position of the factors in f2 and f3 were specified, therefore they did not follow the order of the levels like in f1.


### Matrices and Arrays

- Adding a dim attribute allows it to behave like a multi-dimensional array
- A special case of the multi-dimensional array is the matrix - Only 2 dimensions

Three ways to create an array:
```{r}
matrix(1:6, nrow=3, ncol=2)


array(1:12, c(2,3,2))

c <- 1:6

dim(c) <- c(3,2)

c
```

length() -> nrow() and ncol() for matrices, dim() for arrays
names() -> rownames() and colnames() for matrices, dimnames() for arrays
c() -> cbind() and rbind()

Note: It is possible to have a single dimension matrix
```{r}
matrix(1:3, nrow=1)
# str() reveals the differences.
str(1:3)
str(matrix(1:3, nrow=1))
```

Note: Lists can also be made as matrices and arrays

### Exercises
1. What does dim() return when applied to a vector?
  - dim() returns null, since vectors do not have the attribute dim() applied to them
  
```{r}
dim(1:3)

test_vec <- c(1,3,5)

dim(test_vec)
```

2. If is.matrix(x) is TRUE, what will is.array(x) return?
  - Also true, since a matrix is inherently a special condition of an array
  
```{r}
test_matrix <- matrix(1:6, nrow=3, ncol=2)

is.matrix(test_matrix)

is.array(test_matrix)
```

3. How would you describe the following three objects? What makes them different to 1:5?
```{r}
x1 <- array(1:5, c(1, 1, 5))
x2 <- array(1:5, c(1, 5, 1))
x3 <- array(1:5, c(5, 1, 1))
```

  - x1 is 5 instances a and 1x1 matrix (my way of saying 1x1x5 array)
  - x2 is a matrix with 1 row, 5 columns 
  - x3 is a matrix with 5 rows and 1 column
  
  - These are all inherently matrices, and will have different properties attached to them
    - For example the dim() attribute
    
    
```{r}
print("x1")
print(x1)
print("dimensions of x1")
dim(x1)
print("x2")
print(x2)
print("dimensions of x2")
dim(x2)
print("x3")
print(x3)
print("dimensions of x3")
dim(x3)

print("c(1,2,3,4,5")
print(1:5)
print("will return null")
dim(1:5)

```

## Data frames
- Data frame is the most common way of storing data in R
- Data frame is a list of equal length vectors -> 2 dimensional

- Shares properties of both matrix and list:
  - has names(), colnames() [These are equivalent], rownames()
  - length() is the length of the underlying list. equivalent to ncols()
  - nrow() gives number of rows
  - Can be subsetted as a 1D array or 2D list.
  
### Creating Dataframes
- Takes named vectors as input
```{r}
df <- data.frame(x = 1:3, y=c("a", "b", "c"))

nrow(df)
ncol(df)
names(df)
str(df)
df


```
Notice that y is evaluated as a factor. This may not be good! Can use argument stringsAsFactors = FALSE to fix

### Testing and Coercion
- Type of a dataframe returns list. Need to check class() of dataframe to confirm it is what it looks like
- Can also run is.data.frame()

- Can coerce to dataframe
```{r}
vec <- c(1:10)

a_list <- list(x=1:5, y=c("a","b", "c", "d", "e"))
b_list <- list(x=1:5, y=c("a","b", "c"))

m <- matrix(1:10, nrow=5, ncol=2)

# Will return 2 columns: x, y
as.data.frame(a_list)

# Will return a df with mirrored structure
as.data.frame(m)

# Will return error
as.data.frame(b_list)
# ayyyy
```

### Combining Data Frames

- Can combine dataframes using cbind() and rbind() (LIKE MATRICES!)
  - cbind() adds as columns
    - The number of rows must match
  - rbind() adds as rows
    - The number of columns and column names must match
```{r}
cbind(df, data.frame(z = 3:1))
rbind(df, data.frame(x = 10, y = "z"))
```

- Note: Cannot cbind() vectors together to create a dataframe. One element must already be a dataframe for them to be coerced into one.


### Special Columns
- Dataframe is a list of vectors
  - Therefore a column can be a list
  
```{r}
df <- data.frame(x=1:3)
df$y <- list(1:2, 1:3, 1:4)

df

```

- Although this doesn't work during assignment. Must wrap the list in I() first.
- Same applies to a dataframe with a matrix or array
- This is fairly dangerous behaviour, since most functions assume each column is an atomic vector.

### Exercises

1. What attributes does a data frame possess?
  - colnames() (or names())
  - ronames()
  - class()
  - dim()
  
```{r}
df <- data.frame(x=1:5, y=c("a","b","c","d","e"), z=c(TRUE, FALSE, TRUE, TRUE, FALSE))

attributes(df)

dim(df)
```

2. What does as.matrix() do when applied to a data frame with columns of different types?
  - Similar to how vectors coerce to the most flexible datatype, the matrix coerces to the most flexible datatype
```{r}
as.matrix(df)
str(as.matrix(df))
new_df <- df[,c("x","z")]
as.matrix(new_df)
str(as.matrix(new_df))
```


3. Can you have a data frame with 0 rows? What about 0 columns?
  - When subsetting, you can create a dataframe with zero rows
  - You can create a dataframe of an empty list, since a dataframe is merely a list of vectors
```{r}
df2 <- data.frame(x=1:2, y=c("a","b"))
df2[df2$x ==3]


df3 <- data.frame(list())

df3
```

