---
title: "Advanced_R-Subsetting"
author: "Chris Lavoie"
date: "February 26, 2018"
output: html_document
---

## Subsetting

### Overview

- Three subsetting operators
- Six types of subsetting
- There are multiple differences in behaviour for each object type
- Use subsetting with assignment

### Data types

#### Atomic Vectors

```{r}
x <- c(1.1,2.2,3.3,4.4)

# Positive Integers gives location
# Grabs vector position 1 and 3
x[c(1,3)]

# Grabs item 2
x[2]

# Integer indexs are truncated -> Doesn't evaluate the decimal, only the integer
# Will both return position one
x[c(1.9, 1.1)]

# Negative integers omit that selection
# Returns all but the second position
x[-2]

x[-c(1,4)]

# Cannot mix positive and negative

```

#### Atomic Vectors and Logic

- The most important part of subsetting
- Expressions evaluate to logical operations
- Logical evaluations are applied to each element
```{r}

#
x > 2
# Will return all elements where the expression evaluates to true
x[x > 2]
```
- If logical is shorter than the length of the vector, the logical vector will be recyled until it is equal length

- Subsetting NA always yields NA

```{r}
x[c(TRUE, FALSE, TRUE, NA)]
```
- Character vectors can return elements if the vector is named and the names match
  - Names must match exactly!
```{r}
x <- c(a=1,b=2)

x[1]
x["a"]
#TRUE
x["a"] == x[1]
```
- Name and integer indices can be repeated



### Lists
- Work like vectors
- using [] will return a list.
- using [[]] will return a component of a list
- $ is equivalent to [[]]

```{r}
l <- list(a=1, b=c(1,2), c="fork")

# Returns list of 1
str(l[2])
# Returns element
str(l[[3]])

l["c"]
```


### Matrices and Arrays
- Subset via
  - Using multiple vectors
  - Using a single vector
  - Using a matrix
  
- Most common way is by supplying a 1d index for each dimension

```{r}
m <- matrix(1:100, nrow=10, ncol=10)

m[1,1]
# 2nd row, 6th column
m[2,6]
# Third row, whole column
m[3,]
# Rows 1 2 and 3, columns 2 3 4 and 5
m[1:3, 2:5]

str(m[m > 50])
```

- Using [] will simplify to the lowest possible dimension
