---
title: "Advanced_R_Functions"
author: "Chris Lavoie"
date: "February 27, 2018"
output: html_document
---

# Functions

- Objects in their own right
  - Can work with them in the exact same way as any other object


## Function Components
- Each function has three parts
  - body() - Code inside the function
  - formals() - List of arguments
  - environment() - the 'map' of the location of the function's variables
  
```{r}
f <- function(x) x^2

body(f)
formals(f)
environment(f)
```

- Assignment of body can also modify functions

- Functions also contain attributes (since they are an object)
  - ex. srcref points to the source code of the function
  
```{r}
attributes(f)
```



## Primitive Functions
- Primitive functions have no components. They are functions written in C
- Only found in the {base} package

### Exercises
1.What function allows you to tell if an object is a function? What function allows you to tell if a function is a primitive function?
```{r}
class(f)
class(sum)
```


2. This code makes a list of all functions in the base package.
Use it to answer the following questions:

a.Which base function has the most arguments?

b.How many base functions have no arguments? What's special about those functions?

c. How could you adapt the code to find all primitive functions?
```{r}
objs <- mget(ls("package:base"), inherits = TRUE)
funs <- Filter(is.function, objs)


length_of_formals <- function(x) return(length(formals(x)))

fun_lens <- lapply(funs, length_of_formals)

which.max(unlist(fun_lens))
```
```{r}

fun_no_formals <- fun_lens[fun_lens == 0]

length(fun_no_formals)

prims <- Filter(is.null, formals(objs))

```


## Lexical Scoping
- Set of rules that govern how R looks up the value of a symbol

Lexical Scoping: Implemented automatically at the language level
  - Looks up symbol values based on how functions were nested when created
    - not when they were called
  
  - Four basic principles
    - Name Masking
      - If a name isn't defined inside a function, the function will look up
        - This goes all the way until it reaches a global variable
        - Same is true for closures (functions inside functions)
    - Functions v. Variables
      - Finding functions works the exact same way as finding variables
        - However, if you are clearly looking for a function (like n()), R will ignore non-functions in its search
        - It doesn't mean it's wise to do so.
    - A fresh Start
      - Every time a function is called, a new environment is created to host the execution
      - Each invocation is a completely different state
    - Dynamic Lookup
      - R looks for values when the function is run, not created
      - This is problematic when functions are not self-contained
        -.... Always make your functions self contained!
        
  
Dynamic Scoping: Used in select functions to save typing during analysis


